---
title: "R Examples"
---

This page provides comprehensive examples of using **unifres** with various model types in R.

## Example 1: Logistic Regression

### Detecting Missing Polynomial Terms

```{r}
#| eval: false
library(unifres)

# Generate data with a quadratic relationship
set.seed(1217)
n <- 1000
x <- rnorm(n)
z <- 1 - 2*x + 3*x^2 + rlogis(n)
y <- ifelse(z > 0, 1, 0)

# Fit incorrect model (missing x^2)
fit_wrong <- glm(y ~ x, family = binomial)

# Fit correct model
fit_correct <- glm(y ~ x + I(x^2), family = binomial)

# Compare with function-function plots
par(mfrow = c(1, 2))
ffplot(fit_wrong, type = "l", main = "Missing x²")
ffplot(fit_correct, type = "l", main = "Correct Model")
```

### Using FRED Plots

```{r}
#| eval: false
library(hexbin)

# FRED plots show where the model fails
par(mfrow = c(1, 2))
fredplot(fit_wrong, x = x, type = "hex",
         main = "Wrong Model", aspect = 1)
fredplot(fit_correct, x = x, type = "hex",
         main = "Correct Model", aspect = 1)
```

The wrong model shows a clear parabolic pattern in the FRED plot, indicating the missing quadratic term.

---

## Example 2: Poisson Regression

### Count Data Analysis

```{r}
#| eval: false
# Simulate count data
set.seed(42)
n <- 500
x1 <- rnorm(n)
x2 <- rnorm(n)
lambda <- exp(1 + 0.5*x1 + 0.8*x2)
y <- rpois(n, lambda = lambda)

# Fit model
fit <- glm(y ~ x1 + x2, family = poisson)

# Check model adequacy
par(mfrow = c(2, 2))
ffplot(fit, type = "l", main = "Function-Function Plot")
fredplot(fit, x = x1, type = "hex", main = "FRED: x1")
fredplot(fit, x = x2, type = "hex", main = "FRED: x2")

# Surrogate residual plot
surr_res <- fresiduals(fit, type = "surrogate")
plot(fitted(fit), surr_res,
     xlab = "Fitted Values", ylab = "Surrogate Residuals")
abline(h = 0.5, col = "red", lty = 2)
```

### Detecting Overdispersion

```{r}
#| eval: false
# Generate overdispersed count data
set.seed(123)
n <- 500
x <- rnorm(n)
mu <- exp(1 + x)
# Add overdispersion via negative binomial
y <- rnegbin(n, mu = mu, theta = 2)

# Fit Poisson (wrong)
fit_pois <- glm(y ~ x, family = poisson)

# Fit Quasi-Poisson (correct)
fit_qpois <- glm(y ~ x, family = quasipoisson)

# Compare
par(mfrow = c(1, 2))
ffplot(fit_pois, main = "Poisson")
ffplot(fit_qpois, main = "Quasi-Poisson")
```

---

## Example 3: Negative Binomial Regression

```{r}
#| eval: false
library(MASS)

# Generate negative binomial data
set.seed(2024)
n <- 400
x <- rnorm(n)
mu <- exp(1 + 0.5*x)
y <- rnbinom(n, mu = mu, size = 3)

# Fit model
fit_nb <- glm.nb(y ~ x)

# Diagnostics
fres <- fresiduals(fit_nb)

par(mfrow = c(1, 3))
ffplot(fit_nb, type = "l", main = "Fn-Fn Plot")
fredplot(fit_nb, x = x, type = "hex", main = "FRED Plot")

# Compare to traditional deviance residuals
dev_res <- residuals(fit_nb, type = "deviance")
surr_res <- fresiduals(fit_nb, type = "surrogate")
plot(surr_res, dev_res,
     xlab = "Surrogate Residuals",
     ylab = "Deviance Residuals")
```

---

## Example 4: Generalized Additive Models (GAM)

```{r}
#| eval: false
library(mgcv)

# Generate data with non-linear relationship
set.seed(777)
n <- 500
x <- runif(n, -3, 3)
f <- function(x) sin(x) + 0.5*x
eta <- f(x)
mu <- exp(eta)
y <- rpois(n, lambda = mu)

# Fit GAM with smooth term
fit_gam <- gam(y ~ s(x), family = poisson)

# Fit incorrect GLM (linear)
fit_glm <- glm(y ~ x, family = poisson)

# Compare diagnostics
par(mfrow = c(2, 2))
ffplot(fit_glm, main = "GLM (Wrong)")
ffplot(fit_gam, main = "GAM (Correct)")
fredplot(fit_glm, x = x, type = "hex", main = "GLM FRED")
fredplot(fit_gam, x = x, type = "hex", main = "GAM FRED")
```

---

## Example 5: Ordinal Regression

```{r}
#| eval: false
library(VGAM)

# Generate ordinal outcome data
set.seed(999)
n <- 500
x1 <- rnorm(n)
x2 <- rnorm(n)

# True latent variable
z <- 1 + 0.8*x1 - 0.5*x2 + rlogis(n)

# Create ordinal outcome with 4 categories
y <- cut(z, breaks = c(-Inf, -1, 0, 1, Inf), labels = FALSE)

# Fit proportional odds model
fit_ord <- vglm(factor(y) ~ x1 + x2,
                family = cumulative(parallel = TRUE))

# Diagnostics
par(mfrow = c(1, 3))
ffplot(fit_ord, type = "l", main = "Fn-Fn Plot")
fredplot(fit_ord, x = x1, type = "hex", main = "FRED: x1")
fredplot(fit_ord, x = x2, type = "hex", main = "FRED: x2")

# Check proportional odds assumption
# If violated, try parallel = FALSE
```

---

## Example 6: Zero-Inflated Models

```{r}
#| eval: false
library(pscl)
library(VGAM)

# Generate zero-inflated count data
set.seed(2025)
n <- 500
x <- rnorm(n)

# Zero-inflation probability
pi <- plogis(-1 + 0.5*x)
# Count process
lambda <- exp(1 + 0.8*x)

# Generate outcome
y <- ifelse(rbinom(n, 1, pi) == 1, 0, rpois(n, lambda))

# Fit zero-inflated Poisson model
fit_zip <- zeroinfl(y ~ x | x, dist = "poisson")

# Diagnostics
par(mfrow = c(1, 2))
ffplot(fit_zip, type = "l", main = "ZIP Model")
fredplot(fit_zip, x = x, type = "hex", main = "FRED Plot")

# Compare to standard Poisson
fit_pois <- glm(y ~ x, family = poisson)
ffplot(fit_pois, main = "Standard Poisson")
```

---

## Example 7: Model Comparison

### Comparing Nested Models

```{r}
#| eval: false
# Generate data
set.seed(333)
n <- 800
x1 <- rnorm(n)
x2 <- rnorm(n)
x3 <- rnorm(n)

# True model includes x1 and x2, but not x3
eta <- 1 + x1 + 0.5*x2
y <- rbinom(n, 1, plogis(eta))

# Fit models
fit1 <- glm(y ~ x1, family = binomial)
fit2 <- glm(y ~ x1 + x2, family = binomial)
fit3 <- glm(y ~ x1 + x2 + x3, family = binomial)

# Visual comparison
par(mfrow = c(2, 3))
ffplot(fit1, main = "Model 1: x1 only")
ffplot(fit2, main = "Model 2: x1 + x2")
ffplot(fit3, main = "Model 3: x1 + x2 + x3")

fredplot(fit1, x = x1, type = "hex", main = "M1 FRED")
fredplot(fit2, x = x1, type = "hex", main = "M2 FRED")
fredplot(fit3, x = x1, type = "hex", main = "M3 FRED")
```

---

## Example 8: Customizing Plots

### Function-Function Plots

```{r}
#| eval: false
# Fit a model
fit <- glm(y ~ x, family = binomial)

# Customize ffplot
ffplot(fit,
       type = "p",              # Points instead of line
       pch = 16,                # Solid circles
       col = "steelblue",       # Custom color
       cex = 0.5,               # Smaller points
       ref.col = "red",         # Red reference line
       ref.lwd = 2,             # Thicker reference line
       ref.lty = "dashed",      # Dashed reference line
       xlab = "Theoretical",
       ylab = "Observed",
       main = "Custom Fn-Fn Plot")
```

### FRED Plots

```{r}
#| eval: false
# Customize fredplot with hex type
library(hexbin)
fredplot(fit, x = x,
         type = "hex",
         xbins = 30,                    # Number of hexagonal bins
         colorkey = TRUE,               # Show legend
         color.palette = heat.colors,   # Custom color scheme
         smooth = TRUE,                 # Add LOESS smooth
         smooth.col = "cyan",           # Cyan smooth line
         smooth.lwd = 3,                # Thick smooth line
         xlab = "Predictor",
         ylab = "Functional Residual",
         main = "Custom FRED Plot")

# Customize with KDE type
library(MASS)
library(lattice)
fredplot(fit, x = x,
         type = "kde",
         n = 50,                        # Grid resolution
         color.palette = topo.colors,   # Custom colors
         smooth = TRUE,
         xlab = "Predictor",
         main = "KDE FRED Plot")
```

---

## Example 9: Large Datasets

### Subsampling for Speed

```{r}
#| eval: false
# Large dataset
set.seed(12345)
n <- 50000
x <- rnorm(n)
y <- rbinom(n, 1, plogis(x))
fit <- glm(y ~ x, family = binomial)

# Use subsampling for faster plotting
ffplot(fit, n = 1000)  # Use only 1000 observations

# For FRED plots, sample the data
idx <- sample(n, 5000)
fredplot(fit, x = x[idx], type = "hex")
```

---

## Example 10: Extracting Residuals for Custom Analysis

```{r}
#| eval: false
# Fit model
fit <- glm(y ~ x, family = poisson)

# Get functional residuals
fres_list <- fresiduals(fit, type = "function")

# Evaluate first functional residual at specific points
t_vals <- seq(0, 1, by = 0.1)
f1_vals <- sapply(t_vals, fres_list[[1]])

# Get surrogate residuals for custom plots
surr_res <- fresiduals(fit, type = "surrogate")

# Custom analysis
library(ggplot2)
df <- data.frame(x = x, residual = surr_res)
ggplot(df, aes(x = x, y = residual)) +
  geom_point(alpha = 0.3) +
  geom_smooth(method = "loess", color = "red") +
  geom_hline(yintercept = 0.5, linetype = "dashed") +
  theme_minimal() +
  labs(title = "Custom Surrogate Residual Plot")
```

---

## Tips and Tricks

### 1. Choosing Between Hex and KDE

- **Hex plots**: Better for large datasets, clearer for discrete patterns
- **KDE plots**: Smoother appearance, better for continuous relationships
- Start with hex for exploratory analysis

### 2. Interpreting FRED Plots

- **Horizontal band**: Good model fit
- **Curvature**: Missing polynomial terms
- **Funneling**: Heteroscedasticity
- **Gaps**: Zero-inflation or missing categories

### 3. When to Use Each Residual Type

- **function**: Full diagnostic capabilities, use for formal tests
- **surrogate**: Quick checks, traditional residual plots
- **probscale**: When you want residuals centered at 0

### 4. Combining with Other Diagnostics

```{r}
#| eval: false
# Comprehensive diagnostic panel
par(mfrow = c(2, 3))

# Functional residual diagnostics
ffplot(fit, main = "Fn-Fn Plot")
fredplot(fit, x = x, type = "hex", main = "FRED Plot")

# Traditional diagnostics
plot(fit, which = 1)  # Residuals vs Fitted
plot(fit, which = 2)  # Q-Q Plot
plot(fit, which = 3)  # Scale-Location
plot(fit, which = 5)  # Residuals vs Leverage
```

---

## Next Steps

- Explore [Python Examples](examples-python.qmd) for Python-specific implementations
- Learn about the [Methodology](methodology.qmd) behind functional residuals
- Check the [R API Reference](reference_r/index.qmd) for detailed function documentation
- Read the [Contributing Guide](contributing.qmd) to contribute new examples
