---
title: "Quick Start"
execute:
  output: true
  cache: true
---

```{r}
#| include: false
#| cache: false
# Ensure reticulate uses the correct Python environment
library(reticulate)
# Use local venv if it exists, otherwise use system Python
local_python <- "python/.venv/bin/python"
if (file.exists(file.path("..", local_python))) {
  use_python(file.path("..", local_python), required = TRUE)
} else if (nzchar(Sys.getenv("RETICULATE_PYTHON"))) {
  # Use environment variable if set (for CI/CD)
  use_python(Sys.getenv("RETICULATE_PYTHON"), required = TRUE)
} else {
  # Auto-detect system Python
  use_python(Sys.which("python3"), required = FALSE)
}
```

This guide will get you up and running with **unifres** in just a few minutes.

## R quick start

### Basic example: logistic regression

```{r}
#| warning: false
#| message: false
#| fig-width: 8
#| fig-height: 5
library(unifres)

# Generate example data with a quadratic relationship
set.seed(1217)
n <- 1000
x <- rnorm(n)
z <- 1 - 2*x + 3*x^2 + rlogis(n)
y <- ifelse(z > 0, 1, 0)

# Fit two models: one wrong (missing quadratic), one correct
fit_wrong <- glm(y ~ x, family = binomial)
fit_correct <- glm(y ~ x + I(x^2), family = binomial)

# Compute functional residuals
fres_wrong <- fresiduals(fit_wrong)
fres_correct <- fresiduals(fit_correct)

# Create function-function plots to assess model fit
par(mfrow = c(1, 2))
ffplot(fit_wrong, main = "Wrong Model", type = "l")
ffplot(fit_correct, main = "Correct Model", type = "l")
```

The function-function plot should show deviation from the diagonal for the wrong model and close adherence for the correct model.

### Using different residual types

```{r}
#| warning: false
#| message: false
#| fig-width: 8
#| fig-height: 5
# Functional residuals (list of functions)
fres_fun <- fresiduals(fit_correct, type = "function")
plot(fres_fun[[1]])  # Plot first functional residual

# Surrogate residuals (random samples)
fres_surr <- fresiduals(fit_correct, type = "surrogate")
plot(x, fres_surr)
abline(h = 0.5, col = "red", lty = 2)

# Probability-scale residuals
fres_prob <- fresiduals(fit_correct, type = "probscale")
plot(x, fres_prob)
abline(h = 0, col = "red", lty = 2)
```

### Functional residual density plots

```{r}
#| warning: false
#| message: false
#| fig-width: 8
#| fig-height: 5
library(hexbin)

# FRED plots help identify missing covariates or functional forms
par(mfrow = c(1, 2))
fredplot(fit_wrong, x = x, type = "hex", main = "Wrong Model")
fredplot(fit_correct, x = x, type = "hex", main = "Correct Model")
```

The wrong model should show patterns in the FRED plot, while the correct model should show random scatter.

---

## Python quick start

```{python}
#| include: false
#import matplotlib
#matplotlib.use('Agg')
```

### Basic example: logistic regression

```{python}
#| warning: false
import numpy as np
import statsmodels.api as sm
from unifres import fresiduals, ffplot, fredplot
import matplotlib.pyplot as plt

# Generate example data with a quadratic relationship
np.random.seed(1217)
n = 1000
x = np.random.normal(size=n)
z = 1 - 2*x + 3*x**2 + np.random.logistic(size=n)
y = np.where(z > 0, 1, 0)

# Fit two models: one wrong (missing quadratic), one correct
X_wrong = sm.add_constant(x)
X_correct = sm.add_constant(np.column_stack([x, x**2]))

model_wrong = sm.GLM(y, X_wrong, family=sm.families.Binomial()).fit()
model_correct = sm.GLM(y, X_correct, family=sm.families.Binomial()).fit()

# Create function-function plots
fig, axes = plt.subplots(1, 2, figsize=(12, 5))
ffplot(model_wrong, ax=axes[0])
axes[0].set_title("Wrong Model")
ffplot(model_correct, ax=axes[1])
axes[1].set_title("Correct Model")
plt.tight_layout()
plt.show()
```

### Using different residual types

```{python}
#| warning: false
# Functional residuals (list of scipy distribution objects)
fres_fun = fresiduals(model_correct, type="function")
t_vals = np.linspace(0, 1, 100)
plt.plot(t_vals, fres_fun[0].cdf(t_vals))
plt.xlabel("t")
plt.ylabel("F(t)")
plt.title("First Functional Residual")
plt.show()

# Surrogate residuals (random samples)
fres_surr = fresiduals(model_correct, type="surrogate")
plt.scatter(x, fres_surr, alpha=0.5)
plt.axhline(y=0.5, color='r', linestyle='--')
plt.xlabel("x")
plt.ylabel("Surrogate Residual")
plt.show()

# Probability-scale residuals
fres_prob = fresiduals(model_correct, type="probscale")
plt.scatter(x, fres_prob, alpha=0.5)
plt.axhline(y=0, color='r', linestyle='--')
plt.xlabel("x")
plt.ylabel("Probability-scale Residual")
plt.show()
```

### Functional residual density plots

```{python}
#| warning: false
# FRED plots help identify missing covariates or functional forms
fig, axes = plt.subplots(1, 2, figsize=(12, 5))

fredplot(model_wrong, x, type="kde", ax=axes[0])
axes[0].set_title("Wrong Model")

fredplot(model_correct, x, type="kde", ax=axes[1])
axes[1].set_title("Correct Model")

plt.tight_layout()
plt.show()
```

---

## Understanding the output

### Function-Function (Fn-Fn) Plots

The `ffplot()` function creates a plot of the average functional residual against the theoretical CDF of a Uniform(0,1) distribution.

**Interpretation:**
- **Good fit**: Points follow the 45-degree line closely
- **Poor fit**: Systematic deviations from the diagonal indicate model misspecification
- This is analogous to a Q-Q plot but for functional residuals

### Functional Residual Density (FRED) Plots

The `fredplot()` function visualizes the density of functional residuals against a covariate.

**Interpretation:**
- **Good fit**: Random scatter with no patterns
- **Poor fit**: Systematic patterns suggest:
  - Missing covariates
  - Incorrect functional form (e.g., missing polynomial terms)
  - Incorrect link function

### Residual types comparison

| Type | R | Python | Use Case |
|------|---|--------|----------|
| **function** | List of R functions | List of scipy distributions | Full diagnostic information |
| **surrogate** | Numeric vector | NumPy array | Traditional residual plots |
| **probscale** | Numeric vector | NumPy array | Centered around 0 |

---

## Supported models

### R

- `glm()` - Binomial, Poisson, Quasi-Poisson
- `MASS::glm.nb()` - Negative Binomial
- `mgcv::gam()` - Generalized Additive Models
- `VGAM::vglm()` - Ordinal regression
- `pscl::zeroinfl()` - Zero-inflated Poisson

### Python

- `statsmodels.GLM` - Binomial, Poisson, Negative Binomial
- `statsmodels.NegativeBinomial` - Negative Binomial (discrete)
- `statsmodels.OrderedModel` - Ordinal regression

---

## Next steps

- Explore [R Examples](examples-r.qmd) for detailed use cases
- Explore [Python Examples](examples-python.ipynb) for detailed use cases
- Read about the [Methodology](methodology.qmd) behind functional residuals
- Check the API reference: [R](reference_r/index.qmd) | [Python](reference_python/index.qmd)

---

## Common patterns

### Checking for missing quadratic terms

```{r}
#| warning: false
#| message: false
#| fig-width: 8
#| fig-height: 5
# Generate data
set.seed(1217)
x <- rnorm(100)
y <- rbinom(100, 1, plogis(x + x^2))
fit <- glm(y ~ x, family = binomial)

# If FRED plot shows a parabolic pattern, try adding x^2
fredplot(fit, x = x)  # Shows parabola?
fit2 <- update(fit, . ~ . + I(x^2))
fredplot(fit2, x = x)  # Better?
```

### Comparing multiple models

```{r}
#| warning: false
#| message: false
#| fig-width: 8
#| fig-height: 8
# Generate data and fit two models
set.seed(42)
x <- rnorm(100)
y <- rbinom(100, 1, plogis(0.5 + 0.8*x))
model1 <- glm(y ~ 1, family = binomial)
model2 <- glm(y ~ x, family = binomial)

# Compare models side by side
par(mfrow = c(2, 2))
ffplot(model1, main = "Model 1")
ffplot(model2, main = "Model 2")
fredplot(model1, x = x, main = "Model 1")
fredplot(model2, x = x, main = "Model 2")
```

### Working with large datasets

```{r}
#| warning: false
#| message: false
# Large dataset
set.seed(42)
n <- 2000
x <- rnorm(n)
y <- rpois(n, exp(0.5 + 0.3*x))
fit <- glm(y ~ x, family = poisson)

# Subsample for faster plotting
ffplot(fit, n = 1000)  # Use 1000 random observations
fredplot(fit, x = x[1:1000])  # Plot first 1000 points
```
